---
title: "Practical Fall 2019"
author: "Julie Heng"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Jules/Documents/UMich_Bio201_F19/")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")
```

- Character, because of the quotation marks and word three typed out

### Question 2

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]

- Inner joins only keep the rows that match in both data frames according to specified columns. So extra rows are dropped.

### Question 3

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

- The three assumptions for a t-test are large sample size (n > 30), normality (Shapiro test and visual check through roughly bell-shaped histogram), and equal variances (var.test)
- If the variance assumption is violated, specify that argument by setting var.equal = FALSE
- If the normality assumption is violated, use a non-parametric test (Wilcox test)

### Question 4

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}
butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

- Result1 will choose the rows with "Fall2019" as the semester. True
- Result2 will keep rows less than or equal to 47 mmol/kg. False
- Result3 will compute the median of Avg_butyrate measurements per semester. False

### Question 5

something about NAs

The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will affect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane))

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

- Option 1 will remove the NA measurements before the methane mean is calculated. This means fewer measurements per participant will be used to calculate a weekly mean.
- Option 2 will keep the NA measurements. This means that there will be fewer Avg_methane data points as a participant's weekly mean cannot be calculated with NAs.

# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
# data import code 
hengj <- read_delim(file = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/Raw_data/hengj.txt",
                          delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>% 
  rename_all(tolower)

breath <- read_delim(file = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/Raw_data/breath.txt",
                          delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>% 
  rename_all(tolower) %>% 
  rename(ch4_ppm = ch4,
         h2_ppm = h2)

taxa <- read_delim(file = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/Raw_data/taxa.txt",
                          delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>% 
  rename_all(tolower)
```

Update column names in all files match the course conventions. [1 point]
```{r}
#completed in previous step
```

View the supplement types in your semester. Are they all entered in the same style? If not update them to match. [1 point]
```{r}
hengj[["supplement_consumed"]] <- recode(.x=hengj[["supplement_consumed"]], 
                                  "accessible"="Accessible")
```

### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(hengj)
#472 by 9

dim(breath)
#2668 by 5

dim(taxa)
#46881 by 6
```

Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
#while a full join will drop as few data points as possible, it doesn't make sense to join all the breath data with one semester's worth of data

joined_df <- inner_join(hengj, breath, by = c("participant_id", "study_week", "semester"))

```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(joined_df)
#483 by 11
```

### Question 8

Calculate the total SCFAs. [1 point]
```{r}
total_scfa <- joined_df %>% 
  mutate(total_scfas_mmol_kg = (acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg))
```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
```{r}
scfa_wkly_means <- total_scfa %>% 
  group_by(participant_id, study_week, supplement_consumed, semester) %>%
  summarize(avg_acetate_mmol_kg = mean(acetate_mmol_kg), 
            avg_butyrate_mmol_kg = mean(butyrate_mmol_kg),
            avg_propionate_mmol_kg = mean(propionate_mmol_kg),
            total_avg_scfas_mmol_kg = mean(total_scfas_mmol_kg),
            avg_h2_ppm = mean(h2_ppm),
            avg_ch4_ppm = mean(ch4_ppm))
```

What are the dimensions of the data frame with the avearges? Write the results as a comment after the function.
```{r}
dim(scfa_wkly_means)
#107 by 10
```

Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data <- inner_join(scfa_wkly_means, taxa, by = c("participant_id", "study_week"))

dim(all_data)
#2580 by 14
```

Save all data frames generated to your GitHub page. [1 point]
```{r}
write_delim(joined_df, 
            path="Practial/curated_data/joined_df.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
write_delim(scfa_wkly_means, 
            path="Practial/curated_data/scfa_wkly_means.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
write_delim(total_scfa, 
            path="Practial/curated_data/total_scfa.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
write_delim(all_data, 
            path="Practial/curated_data/all_data.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
#Because of different y axes variables, start with three individual plots. No legend or x-axis label included for aesthetic clarity.

acetate_plot <- scfa_wkly_means %>% 
  filter(supplement_consumed == "BRMPS+Accessible",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = avg_acetate_mmol_kg)) +
  geom_point(aes(color = study_week)) +
      labs(x = NULL,
           y = "Average acetate (mmol/kg)") +
    theme(legend.position = "none",
          plot.margin = margin(1, .5, .5, .5, "cm"))
butyrate_plot <- scfa_wkly_means %>% 
  filter(supplement_consumed == "BRMPS+Accessible",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = avg_butyrate_mmol_kg)) +
  geom_point(aes(color = study_week)) +
      labs(x = NULL,
           y = "Average butyrate (mmol/kg)") +
    theme(legend.position = "none",
          plot.margin = margin(1, .5, .5, .5, "cm"))
propionate_plot <- scfa_wkly_means %>% 
  filter(supplement_consumed == "BRMPS+Accessible",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = avg_propionate_mmol_kg)) +
  geom_point(aes(color = study_week)) +
      labs(x = NULL,
           y = "Average propionate (mmol/kg)") +
    theme(legend.position = "none",
          plot.margin = margin(1, .5, .5, .5, "cm"))
scfa_wkly_accessible <- plot_grid(acetate_plot, butyrate_plot, propionate_plot, 
          nrow = 1, ncol = 3)
scfa_wkly_accessible_plot <- scfa_wkly_accessible + draw_plot_label(label = "Average SCFA change after BRMPS+Accessible supplement consumption, Fall 2016", fontface = "bold", size = 14)
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/scfa_wkly_accessible_plot.pdf",
          plot = scfa_wkly_accessible_plot,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

#violin plot
acetate_violin_plot <- scfa_wkly_means %>%
  filter(supplement_consumed == "BRMPS+Accessible",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = avg_acetate_mmol_kg)) + 
  geom_violin(aes(color = study_week)) +
  labs(x = NULL,
       y = "Average acetate (mmol/kg)") +
    theme(legend.position = "none",
          plot.margin = margin(1, .5, .5, .5, "cm"))
butyrate_violin_plot <- scfa_wkly_means %>%
  filter(supplement_consumed == "BRMPS+Accessible",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = avg_butyrate_mmol_kg)) + 
  geom_violin(aes(color = study_week)) +
  labs(x = NULL,
       y = "Average butyrate (mmol/kg)") +
    theme(legend.position = "none",
          plot.margin = margin(1, .5, .5, .5, "cm"))
propionate_violin_plot <- scfa_wkly_means %>%
  filter(supplement_consumed == "BRMPS+Accessible",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = avg_propionate_mmol_kg)) + 
  geom_violin(aes(color = study_week)) +
  labs(x = NULL,
       y = "Average propionate (mmol/kg)") +
    theme(legend.position = "none",
          plot.margin = margin(1, .5, .5, .5, "cm"))
scfa_accessible_violin <- plot_grid(acetate_violin_plot, butyrate_violin_plot, propionate_violin_plot, 
          nrow = 1, ncol = 3)
scfa_accessible_violin_plot <- scfa_accessible_violin + draw_plot_label(label = "Average SCFA change after BRMPS+Accessible supplement consumption, Fall 2016", fontface = "bold", size = 14)
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/scfa_accessible_violin_plot.pdf",
          plot = scfa_accessible_violin_plot,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]
```{r}
# t-test assumptions
# check sample size
scfa_wkly_means %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS+Accessible") %>% 
  group_by(study_week) %>% 
  summarise(sample_size = n()) #week1 = 13, week3 = 14 both < 30

# checking for normal distribution - week 1 data
brmps_accessible_df1 <- scfa_wkly_means %>%
  filter(study_week == "week1",
         supplement_consumed == "BRMPS+Accessible")
shapiro.test(accessible_df1$total_avg_scfas_mmol_kg) #p-value = 0.1313 > .1, not significant
brmps_accessible_histogram1 <- ggplot(accessible_df1, aes(x=total_avg_scfas_mmol_kg)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 distribution before BRMPS+Accessible supplement, Fall 2016") #appears non-normal, given skewed data and several outliers
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/brmps_accessible_histogram1.pdf",
          plot = brmps_accessible_histogram1,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

# checking for normal distribution - week 3
brmps_accessible_df3 <- scfa_wkly_means %>%
  filter(study_week == "week3",
         supplement_consumed == "BRMPS+Accessible")
shapiro.test(accessible_df3$total_avg_scfas_mmol_kg) #p-value = 0.05085 < .1, significant
brmps_accessible_histogram3 <- ggplot(accessible_df3, aes(x=total_avg_scfas_mmol_kg)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 distribution after BRMPS+Accessible supplement, Fall 2016") #gaps and outliers
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/brmps_accessible_histogram3.pdf",
          plot = brmps_accessible_histogram3,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

# variances 
var.test(x = brmps_accessible_df1$total_avg_scfas_mmol_kg, 
         y = brmps_accessible_df3$total_avg_scfas_mmol_kg, 
         alternative = "two.sided") #p-value: 0.03516 < .1,  significant - therefore, we assume unequal variances

# non-normal, unequal variances, use wilcoxan test
the_accessible_df <- scfa_wkly_means %>% 
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS+Accessible") %>%
  select(participant_id, study_week, total_avg_scfas_mmol_kg) %>%
  spread(study_week, total_avg_scfas_mmol_kg) 

wilcox.test(x = the_accessible_df$week1, 
       y = the_accessible_df$week3, 
       alternative = "two.sided", paired = TRUE, var.equal = FALSE) #p-value = 0.4688

```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

Given the high p-value, we conclude that there is no change in total average SCFA concentration after BRMPS+Accessible supplement between week 1 and week 3.

### Question 10

Is there a correlation between the abundance of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
h2gen_h2 <- all_data %>%
  filter(category == "hydrogen_gen") %>% 
  group_by(participant_id, study_week, avg_h2_ppm) %>%
  summarize(total_fract = sum(fract))
h2gen_h2_plot <- h2gen_h2 %>% 
  ggplot(aes(x = total_fract,
             y = avg_h2_ppm)) + 
  geom_point() + 
  geom_smooth(method = "lm",
              se = FALSE) +  
  xlab("Hydrogen generators") + 
  ylab("Average breath H2 (ppm)") +
  labs(title = "Breath hydrogen vs. abundance of h2 generators")
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/h2gen_h2_plot.pdf",
          plot = h2gen_h2_plot,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# linear regression test
h2gen_h2_model <- h2gen_h2 %>%
  lm(total_fract ~ avg_h2_ppm, data = .)
summary(h2gen_h2_model)
#p-value: 0.7621
#Adjusted R-value: -0.01887 

```

What can you conclude from these results?

The low (negative!) adjusted-r^2 value and high p-value means this data does not fit a linear model. There is no correlation between the abundance of hydrogen generators and breath hydrogen.

Is there a correlation between the abundance of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
h2con_ch4 <- all_data %>%
  filter(category == "hydrogen_consumers") %>% 
  group_by(participant_id, study_week, avg_ch4_ppm) %>%
  summarize(total_fract = sum(fract))
h2con_ch4_plot <- h2con_ch4 %>% 
  ggplot(aes(x = total_fract,
             y = avg_ch4_ppm)) + 
  geom_point() + 
  geom_smooth(method = "lm",
              se = FALSE) +  
  xlab("Hydrogen consumers") + 
  ylab("Average breath CH4 (ppm)") +
  labs(title = "Breath methane vs. abundance of h2 consumers")
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/h2con_ch4_plot.pdf",
          plot = h2con_ch4_plot,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

h2con_ch4_model <- h2con_ch4 %>%
  lm(total_fract ~ avg_ch4_ppm, data = .)
summary(h2con_ch4_model)
#Adj. R-squared: 0.7984
#p-value: 1.597e-12

```

What can you conclude from these results?

Given the low p-value and relatively high r-squared value (nearly our approximate threshold of .8), this data can be represented by a linear model. There is a roughly linear correlation betweenthe abundance of hydrogen consumers and breath methane.

### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot
h2gen <- taxa %>%
  filter(category == "hydrogen_gen") %>%
  select(participant_id, fract, category)

h2con <- taxa %>%
  filter(category == "hydrogen_consumers") %>%
  select(participant_id, fract, category)

h2gen_h2con <- inner_join(h2gen, h2con, by = c("participant_id")) %>% 
  rename(hydrogen_gen = fract.x,
         hydrogen_consumers = fract.y)
h2gen_h2con_plot <- h2gen_h2con %>% 
  ggplot(aes(x = hydrogen_gen,
             y = hydrogen_consumers)) + 
  geom_point() + 
  geom_smooth(method = "lm",
              se = FALSE) +  
  xlab("Abundance of hydrogen generators") + 
  ylab("Abundance of hydrogen consumers") +
  labs(title = "Abundance of hydrogen consumers vs. generators")
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Practial/figures/h2gen_h2con_plot.pdf",
          plot = h2gen_h2con_plot,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1) #slight negative slope for regression line. More interestingly, pattern has box-like edges

#test
h2gen_h2con_model <- h2gen_h2con %>%
  lm(hydrogen_gen ~ hydrogen_consumers, data = .)
summary(h2gen_h2con_model)
#Adjusted R-squared:  0.02001 
#p-value: 4.837e-07

```

The abundance of hydrogen producers and consumers does not appear to have a linear relationship. However, after seeing the graphical results, it appears that a nonlinear model may better illustrate the relationship. The data is skewed towards (0,0) for both variables.


-----
end