---
title: "Lab 7 - Alpha diversity metrics"
author: "Julie Heng"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Jules/Documents/UMich_Bio201_F19/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
library(phyloseq); packageVersion("phyloseq")
set.seed(7)
```


# Import data

### Sequence variants 
Import shared/ESV/OTU table and format for phyloseq. Phyloseq works by matching rownames across matrix objects (This is similar to how we matched columns in joins of data frames). A two-dimensional matrix in R looks just like a data frame, however R matrices can be three-dimensional. A three-dimensional object is created when multiple two-dimensional objects are combined (we'll work through this in the upcoming sections). 
```{r}
myfunction <- function(x) {
  x = round ( x * 1000 , digits = 0 )
  }

sv_table <- read_delim("C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/raw_data/seq_variants_wide.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  rename(sample_id = X1) %>% #add missing column name 
  mutate(sample_id = str_replace(sample_id, 
                                 pattern = "-", 
                                 replacement = "_")) %>% #replace hyphens with underscores in sample id column 
  mutate_if(is.numeric, myfunction) %>% 
  column_to_rownames(var = "sample_id") %>% #sample IDs need to be made into row names
  as.matrix() %>% #convert data frame to matrix object
  otu_table(., taxa_are_rows = FALSE) #specify type of phyloseq object
```

### Taxa 
Import taxonomy table and format for phyloseq.
```{r}
taxa <- read_delim("C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/raw_data/seqvar_classifications.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  column_to_rownames(var = "sv") %>% #sequence variants (OTUs) need to be made into row names 
  as.matrix() %>%
  tax_table(.)  #specify type of phyloseq object
```

### Participant data 
Import sample data (where a sample is an individual microbiome sample) and format for phyloseq.
```{r}
all_data_indv <- read_delim("C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/raw_data/all_data_indv.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, 
                            col_types = cols(Bristol_numeric = col_number(),
                                             Bristol_score = col_character(), 
                                             pH = col_number())) %>%
  rename_all(tolower) %>% #make all column names lower case
  distinct(., sample_id, .keep_all = TRUE) %>% #remove duplicate sample ids 
  column_to_rownames(var = "sample_id") %>% #sample IDs need to be made into row names
  sample_data(.) #specify type of phyloseq object
```


# Create phyloseq objects
Phyloseq is an R package that has a suite of functions that make microbial community anaylsis a bit simpler. https://joey711.github.io/phyloseq/index.html It requires the data to be structured in three critical files (1) sequence variant (or OTU) table that contains counts of each sequence; (2) Taxanomic information that contains classifications from Phylum to species for each sequence variant; (3) sample information, which is the participant information we have been working with so far this semester. We imported these three files in the section above, now we have to combine them into one object. 
```{r}
physq_obj <- phyloseq(sv_table, taxa, all_data_indv) #combine three phyloseq objects created above 

physq_obj #view dimentions of new object 
```


# Richness
There are numerous alpha diversity metrics, however we are going to focus on number richness, evenness, and Simpson for our analyses. We will calculate, plot, extract information about alpha diversity with the plot_richness() function. 

### Calculate richness
```{r}
samp_data <- read_delim("C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/raw_data/all_data_indv.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, 
                            col_types = cols(Bristol_numeric = col_number(),
                                             Bristol_score = col_character(), 
                                             pH = col_number())) %>% #import sample data again
  rename_all(tolower) %>% #make all column names lower case
  distinct(., sample_id, .keep_all = TRUE) %>%
  select(participant_id, sample_number, sample_id, study_week, 
         semester, supplement_consumed, frequency) #drop extra columns 

# Calculate richness
richness_df <- physq_obj %>%
  subset_samples(., semester != "Winter2015") %>% #subset for semesters of interest
  subset_samples(., study_week == "week1" | study_week == "week3") %>% #subset for weeks of interest
  estimate_richness(., split = TRUE,  measures = c("Observed")) %>% #calculate richness
  rownames_to_column(var = "sample_id") %>% #reverse phyloseq object formatting
  inner_join(samp_data, by = "sample_id") %>% #join with sample data imported above 
  rename(richness = Observed) %>%
  group_by(participant_id, study_week, semester, supplement_consumed) %>%
  summarise(avg_richness = round(mean(richness), digits = 0)) #calculate average richness per person, per week 

# save richness data 
write_delim(richness_df, 
            path="Lab7/curated_data/richness_df.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
```

### Homework 7.1 

We can use these alpha diversity metrics (richness, shannon, or others) in the same manner as any other measured characteristic. So, if we want to determine if the consumption of a supplement changed richness we would use the data frame created in the section above to conduct a paired t-test. 
```{r}
# check assumptions 
# check sample size
richness_df %>%
  filter(study_week == "week1" | study_week == "week3") %>% 
  group_by(study_week) %>% 
  summarise(sample_size = n()) #week1 = 97, week3 = 88 both > 30

# checking for normal distribution - week 1 data
richness_df1 <- richness_df %>%
  filter(study_week == "week1")
shapiro.test(richness_df1$avg_richness) #p-value = 0.9016 > .1, not significant
richness_histogram1 <- ggplot(richness_df1, aes(x=avg_richness)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 richness before BRMPS supplement") #appears normal, bell shape is evident
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/figures/richness_histogram1.pdf",
          plot = richness_histogram1,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

# checking for normal distribution - week 3
richness_df3 <- richness_df %>%
  filter(study_week == "week3")
shapiro.test(richness_df3$avg_richness) #p-value = 0.9783 > .1, not significant
richness_histogram3 <- ggplot(richness_df3, aes(x=avg_richness)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 richness before BRMPS supplement") #appears normal
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/figures/richness_histogram3.pdf",
          plot = richness_histogram3,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

# variances 
var.test(x = richness_df1$avg_richness, 
         y = richness_df3$avg_richness, 
         alternative = "two.sided") #p-value: 0.4641 > .1,  not significant - therefore, we assume equal variances

#We proceed with a paired t-test (parametric), with equal variances
brmps_richness_df <- richness_df %>% 
  filter(study_week == "week1" | study_week == "week3") %>%
  spread(study_week, avg_richness) 
write_delim(brmps_richness_df, 
            path="C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/curated_data/brmps_richness_df.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

# write code for paired t-test here 
t.test(x = brmps_richness_df$week1, 
       y = brmps_richness_df$week3, 
       alternative = "two.sided", paired = TRUE, var.equal = TRUE)

# p-value = 6.977e-07, statistically significant
```

Did you conclude the richness changed from week 1 to week 3 in individuals who consumed BRMPS?
Yes, the extremely small p-value indicates that the richness changed from week 1 to week 3 in individuals who comsumed BRMPS.

### Plot richness

```{r}
# Plot richness
rich1 <- physq_obj %>%
  subset_samples(., semester != "Winter2015") %>% 
  subset_samples(., study_week == "week1" | study_week == "week3") %>%
  plot_richness(., "study_week", measures = c("Observed")) +
  facet_grid("semester") +
  ylab("Richness (Observed ESVs)") + xlab(NULL)
rich1
```

This plot is pretty basic. Below we will update the plot with some of our preffered display settings. 
```{r}
rich1$layers <- rich1$layers[-1] #drop geom_point 

rich2 <- rich1 +   
  geom_violin(aes(color = study_week)) + #add violin in color
  geom_jitter(aes(color = study_week)) +  #add individual points in color 
  theme(legend.position = "none")
rich2

save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/figures/rich2_plot.pdf",
          plot = rich2,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1) #upload plot to your GitHub fork 
```


# Simpson 

To calculate Simpson diveristy index we repeat the process above for calcuating richness, the only thing we have to change is the `measures = c("Simpson")` argument for the plot_richness() and estimate_richness() functions. 

### Calculate simpson
```{r}
simpson_df <- physq_obj %>%
  subset_samples(., semester != "Winter2015") %>% #subset for semesters of interest
  subset_samples(., study_week == "week1" | study_week == "week3") %>% #subset for weeks of interest
  estimate_richness(., split = TRUE,  measures = c("Simpson")) %>% #calculate Simpson Diversity
  rownames_to_column(var = "sample_id") %>% #reverse phyloseq object formatting
  inner_join(samp_data, by = "sample_id") %>% #join with sample data imported above 
  rename(simpson = Simpson) %>%
  group_by(participant_id, study_week, semester, supplement_consumed) %>%
  #calculate average Simpson Diversity per person, per week
  summarise(avg_richness = round(mean(simpson), digits = 2))  #use 2 digits since 0-1 scale 

# save Simpson data 
write_delim(simpson_df, 
            path="Lab7/curated_data/simpson_df.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
```

### Homework 7.2 
Determine if Simpson Diversity differed significantly before vs. during supplement consumption. Write your conclusions in the space below.
```{r}
# check assumptions 
# check sample size
simpson_df %>%
  filter(study_week == "week1" | study_week == "week3") %>% 
  group_by(study_week) %>% 
  summarise(sample_size = n()) #week1 = 97, week3 = 88 both > 30

# checking for normal distribution - week 1 data
simpson_df1 <- simpson_df %>%
  filter(study_week == "week1")
shapiro.test(simpson_df1$avg_richness) #p-value = 1.619e-10 < .1,  significant, not normal
simpson_histogram1 <- ggplot(simpson_df1, aes(x=avg_richness)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 Simpson Diversity Index before BRMPS supplement") #appears skewed left, with several outliers far to the left
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/figures/simpson_histogram1.pdf",
          plot = simpson_histogram1,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

# checking for normal distribution - week 3
simpson_df3 <- simpson_df %>%
  filter(study_week == "week3")
shapiro.test(simpson_df3$avg_richness) #p-value = 9.946e-12 < .1,  significant, not normal
simpson_histogram3 <- ggplot(simpson_df3, aes(x=avg_richness)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 Simpson Diversity Index before BRMPS supplement") #also skewed to the left, with one extreme outlier
save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/figures/simpson_histogram3.pdf",
          plot = simpson_histogram3,
          nrow = 1, ncol = 3,
          base_aspect_ratio = 1.1)

# variances 
var.test(x = simpson_df1$avg_richness, 
         y = simpson_df3$avg_richness, 
         alternative = "two.sided") #p-value: 0.9388 > .1,  not significant - therefore, we assume equal variances

#We proceed with a paired t-test (non-parametric), with equal variances
brmps_simpson_df <- simpson_df %>% 
  filter(study_week == "week1" | study_week == "week3") %>%
  spread(study_week, avg_richness) 
write_delim(brmps_simpson_df, 
            path="C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/curated_data/brmps_simpson_df.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)

# write code for paired t-test here 
wilcox.test(x = brmps_simpson_df$week1, 
       y = brmps_simpson_df$week3, 
       alternative = "two.sided", paired = TRUE, var.equal = TRUE)

# p-value = 0.02852, statistically significant. Simpson Diversity differed significantly from week 1 to week 3 BRMPS consumption.
```

### Plot simpson 

```{r}
# Simpson 
simp1 <- physq_obj %>%
  subset_samples(., semester != "Winter2015") %>% 
  subset_samples(., study_week == "week1" | study_week == "week3") %>%
  plot_richness(., "study_week", measures = c("Simpson")) +
  facet_grid("semester") +
  ylab("Simpson 1-D") + xlab(NULL)

simp1$layers <- simp1$layers[-1] #drop geom_point 

simp2 <- simp1 +   
  geom_violin(aes(color = study_week)) + #add violin in color
  geom_jitter(aes(color = study_week)) +  #add individual points in color 
  theme(legend.position = "none")
simp2

save_plot(filename = "C:/Users/Jules/Documents/UMich_Bio201_F19/Lab7/figures/simp2_plot.pdf",
          plot = simp2,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1) 
```


-----
end